Correcting for bias in high throughput sequencing data

### 2011-06-17 seqbias-paper.tex
paper: finished

\section{Motivation:}
Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
often conflated with protocal-specific bias, influenced by PCR amplification,
and differing primer affinities and mixtures, for example. The result is
potentially inaccurate quantification, affecting de novo gene annotation and
isoform quantification in RNA-Seq, and peak calling in ChiP-Seq.

\section{Results:}
We present a method to measure and correct for these influences using a simple
graphical model. Our model does not rely on existing gene annotations, making it
applicable to any high throughput sequencing data. We evaluate our method on
several datasets, and by multiple criteria, demonstrating that it effectively
decreases bias and increases uniformity.

\section{Availability:}
The method is implemented in the \texttt{seqbias} R/Bioconductor package,
available freely under the LGPL licence from
\href{http://bioconductor.org}{http://bioconductor.org}.

@@

A significant example is the effect of transcript length on quantification.  As
noted by \citet{Oshlack2009} transcript length biases discovery of
differentially expression, since more reads are aligned to longer transcripts,
granting higher statistical significance in tests differential expression.
Recently, some methods have been developed attempting to account for this effect
in gene set enrichment tests \citep{Young2010,Gao2011}.

%% TODO: do something with this sentance.
Additionally,
differing transcript lengths introduce ``edge effects'', as very short fragments
from the end of a transcript will not be sequenced. NEUMA \cite{Lee2010} and
FPKM \cite{Trapnell2010} are two methods that attempt to take this into account.

@@

\begin{figure*}
\centerline{\includegraphics[width=0.8\textwidth]{fig/freqs.eps}}
\caption{Nucleotide frequencies are plotted relative to the start (labeled,
position 0) of each mapped read, respecting strand. The sequence is taken frmo
the genomic context surrounding the read, so that -40 to -1, for example, fall
outside the read sequence itself. The symmetrized Killback-Leibler divergence is
used to summarize the difference in nucleotide frequency compared to a fixed
estimate of background necleotide frequencies made by sampling many positions
nearby mapped reads.  Under the assumption that reads are sampled uniformly from
transcripts, each of the plots should be essentially flat.}
\label{fig:freqs}
\end{figure*}

@@

Lastly, \citet{Roberts2011} have very recently published a description of
another approach, in which sequence probabilities are modeled by variable-order
Markov chains. The structure of these Markov chains are hard-coded, chosen in
advance using a hill-climbing algorithm on a representative dataset. This method
is implemented in the latest version of Cufflinks \citep{Trapnell2010}, and
tightly incorporated into its estimation of transcript abundance.

@@

%% TODO: where should this figure be cited?
\begin{figure}
\centerline{\includegraphics[width=0.35\textwidth]{fig/overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is classifier is trained to descriminate between the set of
sampled foreground and background sequences, (d) and the classifier is evaluated
at each position within a locus, predicting bias.}
\label{fig:overview}
\end{figure}

@@

Towards that end, we propose training a Bayesian network classifier trained on
examples of foreground and background sequences. Though our goal is not
classification, by using the machinery of classification we can avoid a model
that is over-parametrized. Parameters that are not informative in
discriminating between foreground and background are not included in the model.
The Bayesian network can then used to evaluate sequence probability, and thus
bias, at any genomic position.

However, we have so for ignored one complication: the RNA abundance that we wish
to estimate is not itself independent of the nucleotide sequence. Notably,
exonic DNA tends to be more GC-rich than intergenic DNA. If background sequences
are sampled uniformly from the genome we run the risk of incorrectly adjusting
for biological sequence bias, rather than techncical sequence bias.  To avoid
this, we propose using paired training data. Each foreground training
sequence is paired with a background sequence taken from nearby position
that is likely to have similar abundance and general nucleotide composition.
Alternatively, we could pair foreground samples with background samples from
within the same transcript, but we prefer to avoid dependence on existing
gene annotations as such a model would be limited to RNA-Seq analysis in which
trustworthy and reasonably complete gene annotations are available.

The methods proposed \citet{Hansen2010} and \cite{Roberts2011} also treat bias
correction as a problem of estimating foreground and background sequence
probabilities. They differ primarily in how these sequence probabilities are
estimated. \citet{Li2010} estimate reweighting coefficients ($b_i$, in our
notation) directly, given training data consistig of long, highly-expressed
transcripts.

@@

To obtain background training sequences, we randomly offset the positions from
which the foreground sequences were sampled.  The offset is drawn from a
zero-mean gaussian, and rounded to the nearest integer, away from zero.
By using such a scheme, we attempt to mitigate the effects of biological
sequence bias, sampling positions that are more likely to be biologically
similar.

This procuder produces a training set of $n$ sequences with accompanying labels
$T = \{ (s_1, x_1), (s_2, x_2), \dots, (s_n, x_n) \}$. The label $x_i$ is
binary, indicating classification as background ($x_i = 0$) or foreground ($x_i
= 1$).

@@

As we will be estimating parameters and evaluating the likelihood on the same
set of samples, simply maximizing the likelihood would severely overfit the
training set. We thus penalize model complexity heuristically using the Akaike
information criterion. Where $m$ is the number of parameters needed to specify
the model, we maximize,

$$ \ell' = \ell - m $$

Some benifit would be obtained from a more highly tuned complexity penalty.
However, since the model is trained greedily, additional parameters will be
decreasingly informitive, and increasingly similar between foreground and
background. Adding more parameters will have little effect.  Only when $m$ is
allowed to grow exponentially does the prediction become polluted by small
deviations between thousands of uninformitive parameters.

@@

Figure \ref{fig:models} shows examples of the structure learned when this
procedure is applied to several data sets, using 200,000 reads from each.

@@

Because we cannot observe directly the underlying RNA-abundance, our evaluation
strategy relies on testing two assumptions we make of an ideal, unbiased experiment.
\begin{enumerate}
\item Positional nucleotide frequencies (as in Figure \ref{fig:freqs}), measured
from reads within exons, should not differ greatly from frequencies measured by
sampling uniformly within the same exons.
\item Read counts across a single exon should follow, approxmiately, a Poisson
process.
\end{enumerate}

We applied our procedure (labeled ``BN'') as well as those as well as those of
\citet{Li2010} (``GLM'' and ``MART'') and \citet{Hansen2010} (``7mer''), which
are implemented in the R packages \texttt{mseg} and \texttt{Genominator},
respectively. Thhe procedures were applied to four publically available data
sets \citep{Bullard2010, Mortazavi2008, Trapnell2010, Wetterbom2010}, as well as
an unpublished data sets of our own (Table \ref{tab:datasets}).

\begin{table}
\processtable{Datasets on which the methods are evaluated.\label{tab:datasets}}
{
\begin{tabular}{lllr}\toprule
Experiment & Species & Platform & Read Length \\\midrule
\citet{Wetterbom2010} & Chimpanzee & ABI & 33 \\
Katze (unpublished) & Macaque & ABI & 50 \\
\citet{Bullard2010} & Human & Illumina & 35 \\
\citet{Mortazavi2008} & Mouse & Illumina & 33 \\
\citet{Trapnell2010} & Mouse & Illumina & 75 \\\botrule
\end{tabular}
}{}
\end{table}

@@

We evaluated several variations of the heptamer model. The suggested method
involved averaging the frequencies of the first two heptamers of each read. Yet,
we found that in every case, this performed worse than simply counting the
frequencies of the initial heptamer, and thus we report only the latter. The
background frequences are estimated from positions 18--23 in each read.

Our own method was trained on the 200,000 reads from chromosomes 1--8, with the
highest number of duplicates.

@@

%% --------------------

@@

\begin{figure*}
\centerline{\includegraphics[width=0.9\textwidth]{fig/kl.eps}}
\caption{The Kullback-Leibler divergence compares the frequency of $k$-mers
(here, for $k = 1$ and $k = 4$) surrounding the starts of aligned reads to the
frquencies expected under the assumption of unifom sampling from within exons.
A large divergence indicates significant bias. Plotted here is the divergence
from unadjusted reads counts as well as after adjusting reads counts using each
method.}
\label{fig:kl}
\end{figure*}

@@

If the null model has log-likelihood $L$, and the bias-corrected model $L'$, a
simple goodness of fit measure is the improvement in log-likelihood (a statistic
commonly known as McFadden's pesudo-coefficient of determination
\citep{McFadden1974}), defined as,
$$R^2 = 1 - \frac{L'}{L}$$

This measure can be interpreted as the improvement in fit over the null model,
with $R^2 = 1$ indicating a prefect fit, and a negitve numbers, an increasingly
worse fit. A more dubious, yet intuitive interpretation of $R^2$ is the
proportion of variability eplained by the model.

We compue $R^2$ for each of the test exons, giving us a sense of the variability
of the effectiveness of each model. The results of this analysis are plotted in
Figure \ref{fig:pois}.  To summarize each model with a single number, we can
examine the median $R^2$ value, as listed in Table \ref{tab:pois}.

@@

We have withheld any direct comparison to the method described by
\citet{Roberts2011} and implemented in Cufflinks \citep{Trapnell2010}. Though
this method is superficially similar to our own, a proper comparison is
difficult, as the software can not be applied independently of estimating
transcript abundance in FPKM using Cufflinks. Fairness would dictate that
competing methods be substituted in FPKM estimation, or that a seperate
interface be written to the Cufflinks bias correction method---both comparison
requiring significant effort.

Though the Cufflinks method and our own both use graphical models to estimate
sequence probabilities, we make no restriction on the graph other than
acyclicity, and go to considerable effort to efficiently approximate the optimal
structure for each data set, rathor than using a fixed structure, as in
Cufflinks. A ``one size fits all'' approach likely works quite well in many
cases, yet the observed specificity of the bias to protocal and platform argues
against it. For example, the structures learned by our method (Figure
\ref{fig:models}), are considerably different between those sequenced on an
Illumina platform versus an AB platform.

%% TODO: discussion of run-time, etc

Because we do not require annotations, CHiP-Seq, and other high-throughput
sequencing experiments may also benefit from our model. In a preliminary
investigation, we found the sequence bias in one CHiP-Seq experiment
\cite{Cao2010} was less than that observed in any of the RNA-Seq data we
evaluated, however our method is still able effectively correct for the bias
that was observed (see, Supplementary Section 4).
%% TODO: which actual section ?
Protocol differences, as we have seen,
can result in significant differences in observed nucleotide frequencies, so we
may not safely assert that bias in CHiP-Seq data is always low.  Given the weak
assumptions made by our model, our estimation of bias could easily be
incorporated into CHiP-Seq peak calling algorithms, and potentially improve
accuracy.

RNA-Seq is most often used to compare levels of expression, and so a natural
concern is the consistency of the bias between samples. In the data we examined,
the bias appears to be largely, but not entirely consistent (see,
Supplementary Section 5).
%% TODO: which actual section ?
Similarly, in Figure \ref{fig:freqs}, the three data sets
sequenced on the Illumina platform display similar patterns of nonuniformity,
yet differ in magnitude. Batch effects in RNA-Seq remain a legitimate concern,
that should be evaluated and corrected for.
%% TODO: better end sentence

>>>>

\section{Motivation:}
Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
often conflated with protocol-specific sequence bias, influenced by PCR
amplification, and differing primer affinities and mixtures, for example.  The
result is decreased accuracy in many applications, such as de novo gene
annotation and isoform quantification in RNA-Seq, and peak calling in
ChIP-Seq.

\section{Results:}
We present a method to measure and correct for these influences using a simple
graphical model. Our model does not rely on existing gene annotations, making it
applicable to any high throughput sequencing data. We evaluate our method on
several data sets, and by multiple criteria, demonstrating that it effectively
decreases bias and increases uniformity.

\section{Availability:}
The method is implemented in the \texttt{seqbias} R/Bioconductor package,
available freely under the LGPL license from
\href{http://bioconductor.org}{http://bioconductor.org}.

@@

A significant example is the effect of transcript length on quantification.  As
noted by \citet{Oshlack2009} transcript length biases discovery of
differentially expression, since more reads are aligned to longer transcripts,
granting higher statistical significance in tests of differential expression.
Recently, some methods have been developed attempting to account for this effect
in gene set enrichment tests \citep{Young2010,Gao2011}, leading to more accurate
downstream analysis.

@@

\begin{figure*}
\centerline{\includegraphics[width=0.8\textwidth]{fig/freqs.eps}}
\caption{Nucleotide frequencies are plotted relative to the start (labeled,
position 0) of each mapped read, respecting strand. The sequence is taken from
the genomic context surrounding the read, so that -40 to -1, for example, fall
outside the read sequence itself. The symmetrized Kullback-Leibler divergence is
used to summarize the difference in nucleotide frequency compared to a fixed
estimate of background nucleotide frequencies made by sampling many positions
nearby mapped reads.  Under the assumption that reads are sampled uniformly from
transcripts, each of the plots should be essentially flat.}
\label{fig:freqs}
\end{figure*}

@@

Lastly, \citet{Roberts2011} have very recently published a description of
another approach, in which sequence probabilities are modeled by variable-order
Markov chains. The structure of these Markov chains are hard-coded, chosen in
advance using a hill-climbing algorithm on a representative data set. This method
is implemented in the latest version of Cufflinks \citep{Trapnell2010}, and
tightly incorporated into its estimation of transcript abundance.

@@

@@

Towards that end, we propose training a Bayesian network classifier trained on
examples of foreground and background sequences. Though our goal is not
classification, by using the machinery of classification we can avoid a model
that is over-parametrized. Parameters that are not informative in
discriminating between foreground and background are not included in the model.
The Bayesian network can then used to evaluate sequence probability, and thus
bias, at any genomic position. Figure \ref{fig:overview} gives a brief overview
of the proposed model.

\begin{figure}
\centerline{\includegraphics[width=0.35\textwidth]{fig/overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is classifier is trained to discriminate between the set of
sampled foreground and background sequences, (d) and the classifier is evaluated
at each position within a locus, predicting bias.}
\label{fig:overview}
\end{figure}

However, we have so for ignored one complication: the RNA abundance that we wish
to estimate is not itself independent of the nucleotide sequence. Notably,
exonic DNA tends to be more GC-rich than intergenic DNA. If background sequences
are sampled uniformly from the genome we run the risk of incorrectly adjusting
for biological sequence bias, rather than technical sequence bias.  To avoid
this, we propose using paired training data. Each foreground training
sequence is paired with a background sequence taken from nearby position
that is likely to have similar abundance and general nucleotide composition.
Alternatively, we could pair foreground samples with background samples from
within the same transcript, but we prefer to avoid dependence on existing
gene annotations as such a model would be limited to RNA-Seq analysis in which
trustworthy and reasonably complete gene annotations are available.

The methods proposed \citet{Hansen2010} and \cite{Roberts2011} also treat bias
correction as a problem of estimating foreground and background sequence
probabilities. They differ primarily in how these sequence probabilities are
estimated. \citet{Li2010} estimate reweighting coefficients ($b_i$, in our
notation) directly, given training data consisting of long, highly-expressed
transcripts.

@@

To obtain background training sequences, we randomly offset the positions from
which the foreground sequences were sampled.  The offset is drawn from a
zero-mean Gaussian, and rounded to the nearest integer, away from zero.
By using such a scheme, we attempt to mitigate the effects of biological
sequence bias, sampling positions that are more likely to be biologically
similar.

This procedure produces a training set of $n$ sequences with accompanying labels
$T = \{ (s_1, x_1), (s_2, x_2), \dots, (s_n, x_n) \}$. The label $x_i$ is
binary, indicating classification as background ($x_i = 0$) or foreground ($x_i
= 1$).

@@

As we will be estimating parameters and evaluating the likelihood on the same
set of samples, simply maximizing the likelihood would severely overfit the
training set. We thus penalize model complexity heuristically using the Bayesian
Information Criterion \citep{Schwarz1978a}. Where $m$ is the number of
parameters needed to specify the model, we maximize,

$$ \ell' = 2 \ell - m \log n $$

Some benefit would be obtained from a more highly tuned complexity penalty.
However, since the model is trained greedily, additional parameters will be
decreasingly informative, and increasingly similar between foreground and
background. Adding more parameters will have little effect.  Only when $m$ is
allowed to grow exponentially does the prediction become polluted by small
deviations between thousands of uninformative parameters.

@@

Figure \ref{fig:models} shows examples of the structure learned when this
procedure is applied to several data sets, using 100,000 reads from each.

@@

Because we cannot observe directly the underlying RNA-abundance, our evaluation
strategy relies on testing two assumptions we make of an ideal, unbiased experiment.
\begin{enumerate}
\item Positional nucleotide frequencies (as in Figure \ref{fig:freqs}), measured
from reads within exons, should not differ greatly from frequencies measured by
sampling uniformly within the same exons.
\item Read counts across a single exon should follow, approximately, a Poisson
process.
\end{enumerate}

We applied our procedure (labeled ``BN'') as well as those as well as those of
\citet{Li2010} (``GLM'' and ``MART'') and \citet{Hansen2010} (``7mer''), which
are implemented in the R packages \texttt{mseq} and \texttt{Genominator},
respectively. The procedures were applied to four publicly available data
sets \citep{Bullard2010, Mortazavi2008, Trapnell2010, Wetterbom2010}, as well as
an unpublished data sets of our own (Table \ref{tab:datasets}).

\begin{table}
\processtable{Data sets on which the methods are evaluated.\label{tab:datasets}}
{
\begin{tabular}{lllr}\toprule
Experiment & Species & Platform & Read Length \\\midrule
\citet{Wetterbom2010} & Chimpanzee & ABI & 33 \\
Katze (unpublished) & Macaque & ABI & 50 \\
\citet{Bullard2010} & Human & Illumina & 35 \\
\citet{Mortazavi2008} & Mouse & Illumina & 33 \\
\citet{Trapnell2010} & Mouse & Illumina & 75 \\\botrule
\end{tabular}
}{}
\end{table}

@@

We evaluated several variations of the heptamer model. The suggested method
involved averaging the frequencies of the first two heptamers of each read. Yet,
we found that in every case, this performed worse than simply counting the
frequencies of the initial heptamer, and thus we report only the latter. The
background frequencies are estimated from positions 18--23 in each read.

Our own method was trained on the 100,000 reads from chromosomes 1--8, with the
highest number of duplicates.

@@

@@

\begin{figure*}
\centerline{\includegraphics[width=1.00\textwidth]{fig/kl.eps}}
\caption{The Kullback-Leibler divergence compares the frequency of $k$-mers
(here, for $k = 1$ and $k = 4$) surrounding the starts of aligned reads to the
frequencies expected under the assumption of uniform sampling from within exons.
A large divergence indicates significant bias. Plotted here is the divergence
from unadjusted reads counts as well as after adjusting reads counts using each
method.}
\label{fig:kl}
\end{figure*}

@@

If the null model has log-likelihood $L$, and the bias-corrected model $L'$, a
simple goodness of fit measure is the improvement in log-likelihood (a statistic
commonly known as McFadden's pseudo-coefficient of determination
\citep{McFadden1974}), defined as,
$$R^2 = 1 - \frac{L'}{L}$$

This measure can be interpreted as the improvement in fit over the null model,
with $R^2 = 1$ indicating a prefect fit, and a negative numbers, an increasingly
worse fit. A more dubious, yet intuitive interpretation of $R^2$ is the
proportion of variability enplaned by the model.

We compute $R^2$ for each of the test exons, giving us a sense of the variability
of the effectiveness of each model. The results of this analysis are plotted in
Figure \ref{fig:pois}.  To summarize each model with a single number, we can
examine the median $R^2$ value, as listed in Table \ref{tab:pois}.

@@

We have withheld any direct comparison to the method described by
\citet{Roberts2011} and implemented in Cufflinks \citep{Trapnell2010}. Though
this method is superficially similar to our own, a proper comparison is
difficult, as the software can not be applied independently of estimating
transcript abundance in FPKM (fragments per exonic kilobase, per million mapped
reads) using Cufflinks. Fairness would dictate that competing methods be
substituted in FPKM estimation, or that a separate interface be written to the
Cufflinks bias correction method---both comparison requiring significant effort.

Though the Cufflinks method and our own both use graphical models to estimate
sequence probabilities, we make no restriction on the graph other than
acyclicity, and go to considerable effort to efficiently approximate the optimal
structure for each data set, rather than using a fixed structure, as in
Cufflinks. A ``one size fits all'' approach likely works quite well in many
cases, yet the observed specificity of the bias to protocol and platform argues
against it. For example, the structures learned by our method (Figure
\ref{fig:models}), are considerably different between those sequenced on an
Illumina platform versus an AB platform.

Training our model on more reads leads to more accurate estimation of bias, but
increasingly long training time. For example, in our tests, fitting our model to
100,000 reads from the Mortazavi data, training time was approximately 45
minutes, running on one core of a 3Ghz Intel Xeon processor. However, limiting
the training to 25,000 reads leads to a model that is nearly as accurate while
requiring under 4 minutes to train. A full discussion of this trade-off is in
Supplementary Section 6.
%% TODO: what actual section?

Because we do not require annotations, ChIP-Seq, and other high-throughput
sequencing experiments may also benefit from our model. In a preliminary
investigation, we found the sequence bias in one ChIP-Seq experiment
\cite{Cao2010} was less than that observed in any of the RNA-Seq data we
evaluated, however our method is still able effectively correct for the bias
that was observed (see, Supplementary Section 4).
%% TODO: which actual section ?
Protocol differences, as we have seen,
can result in significant differences in observed nucleotide frequencies, so we
may not safely assert that bias in ChIP-Seq data is always low.  Given the weak
assumptions made by our model, our estimation of bias could easily be
incorporated into ChIP-Seq peak calling algorithms, and potentially improve
accuracy.

RNA-Seq is most often used to compare levels of expression, and so a natural
concern is the consistency of the bias between samples. In the data we examined,
the bias appears to be largely, but not entirely consistent (see,
Supplementary Section 5).
%% TODO: which actual section ?
Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced on the
Illumina platform display similar patterns of nonuniformity, yet differ in
magnitude, suggesting that batch effects in RNA-Seq remain a legitimate concern
that should not be dismissed without evaluation.

In summary, we have demonstrated a relatively simple graphical model that
effectively corrects for sequence bias pervasive in RNA-Seq, and to a lesser
extent, ChIP-Seq experiments. In our tests, this model performs at least as
well, and often better than existing methods, and involves fewer requirements or
assumptions. Our model leads to more accurate quantification, and would likely
provide a positive benefit when incorporated into downstream analysis.
